%!TeX root=../thesis.tex

\chapter{Related Work}
\label{cap:related-work}
\label{chap:related-work}

This chapter synthesizes the literature relevant to the empirical evaluation of Hypo-Stage. 
Rather than presenting a standalone systematic review, it draws on the author's earlier exploratory work --- which included a tertiary systematic
mapping study on software architecture for startups and a multivocal literature review
on centralization and decentralization of architectural decision-making --- and
repositions those findings as narrative context for this work. The chapter also
incorporates references from the gray literature. Such sources
are explicitly identified as relevant, despite having weaker evidential
weight than peer-reviewed publications, because they often capture industry trends and real
organizational experience more rapidly than academic publication cycles allow.

% ---------------------------------------------------------------------------
\section{Centralization and Decentralization of Architectural Decisions}
\label{sec:central-decentral}

A recurring theme in both academic and gray literature is the tension between
centralizing architectural authority in a few experienced individuals versus distributing it
across development teams. Organizations pursuing fast flow and team autonomy
increasingly favor decentralized models as observed by ~\cite{salameh2021architecture, deDiana2011arch},
yet several studies report that without explicit governance mechanisms, decentralization
leads to inconsistent decisions and architectural drift as seen in the work of ~\cite{silveiraNeto2025agile, dasanayake2017collaborative, deDiana2011arch}.

The author's preliminary exploratory work examined this tension through two complementary
lenses. Firstly, the author conducted a tertiary systematic mapping study surveying the existing landscape of secondary studies on software architecture for
startup and scale-up organizations, identifying recurring themes in how early-stage
teams approach architecture governance under resource and time constraints. For the second lens, the author 
performed a multivocal literature review that synthesized both academic
publications and gray literature (practitioner blogs, conference talks, and industry
reports), covering three interrelated perspectives: deriving software architecture from
requirements, the emerging socio-technical architecture trend (which situates
architectural decisions within organizational and team structures), and the tension
between centralized and decentralized decision-making.

The multivocal review surfaced a framing where the industry advocates strongly for decentralized
decisions to enable team autonomy and fast flow \cite{skelton2019}, yet the methods and tools available
for architectural decision-making depend heavily on the tacit knowledge of experienced
architects as observed by \cite{souza2019}. Decentralization thus risks transferring decision responsibility to
practitioners who lack the conceptual vocabulary or methodological support to exercise
it effectively. However, the author recognized that this tension reflects a recognized design trade-off 
rather than a genuine logical contradiction, and framing this work around resolving that tension would risk over-extending its scope.
For this reason, this work consequently does not employ that framing; instead, it focuses on the
concrete, tractable problem of providing tool support for one well-defined uncertainty
management technique.

% ---------------------------------------------------------------------------
\section{Socio-Technical Architecture}
\label{sec:sociotechnical}

A significant body of work, both academic and practitioner-oriented, frames
architectural decision-making not merely as a technical concern but as a
\emph{socio-technical} one --- shaped by the organizational structure,
communication patterns, and cognitive constraints of the teams
involved, as observed in ~\cite{tang2017human, razavian2019empirical, herbsleb1999splitting}.
This framing is directly relevant to the motivation for the Hypo-Stage tool,
because tool-embedded scaffolding does not only lower the technical cost of
applying the ArchHypo technique, but it also contributes to changing the social
dynamics of how architectural uncertainty is surfaced and addressed within a
team ~\cite{borowa2023rationales, razavian2019empirical}.

\subsection{Conway's Law and Its Implications}
\label{subsec:conways-law}

The foundational observation underlying the socio-technical framing is Conway's
Law ~\cite{conway1968}: ``Any organization that designs a system (defined broadly) will
produce a design whose structure is a copy of the organization's communication
structure.'' Conway's Law implies that architectural decisions cannot be analyzed in
isolation from the organizational and team structure in which they are made. Considering that perspective, 
this work also supports the idea that team boundaries become architectural boundaries.

~\cite{skelton2019} extended this insight into an actionable framework
with the concept of the \emph{Inverse Conway Maneuver}: rather than allowing the
organization's structure to passively determine the architecture, teams should
deliberately design team topologies to produce the target architecture. In practical
terms, this means that architectural governance --- including how uncertainty is managed
and who is empowered to make decisions --- must be considered a design decision with
organizational as well as technical consequences.

\subsection{Socio-Technical Architecture as an Industry Trend}
\label{subsec:sociotechnical-trend}

The InfoQ Software Architecture and Design Trends Report for 2024 by ~\cite{infoq2024trends}
identified \emph{socio-technical architecture} as a distinct ``Early Adopter'' trend,
explicitly describing it as replacing the earlier framing of ``architecture as a team
sport.'' The report notes that the trend encompasses two related ideas: on one side,
the evolving role of architects as mentors and technical leaders rather than sole
decision-makers; on the other, the growing recognition --- driven in part by the
dissemination of Team Topologies~\cite{skelton2019} --- that the design of a system
must consider all the people who build, support, and maintain it ~\cite{infoq2024trends}.

It is important to note that InfoQ is a practitioner-oriented publication, not a
peer-reviewed academic venue. The Trends Report reflects the editorial judgment of an
experienced editorial board drawing on community submissions, conference talks, and
practitioner case studies. As gray literature, its findings carry weaker evidential
weight than systematic empirical studies; nonetheless, they are valuable as an indicator
of what concepts are gaining traction in industrial practice and what challenges
practitioners are actively facing, as observed in ~\cite{infoq2024trends}.

In a related InfoQ article, ~\cite{infoq2025empowering} described
a transformation journey at a software organization adopting Team Topologies and
decentralized architectural decision-making. Their account illustrates several
socio-technical dynamics that are directly relevant to this work: the initial disbelief
of teams that they are actually empowered to make architectural decisions, the flood of
decisions that follows once empowerment is accepted, the consequences teams face when
decisions have unforeseen cross-boundary impacts, and the eventual self-correction as
teams develop peer accountability mechanisms. All four phases are driven by social
factors at least as much as technical ones.

A further gray-literature source relevant to this work is the InfoQ article by ~\cite{infoq2025decentralized}, which argues that ``distributed architecture is
more social than technical.'' Tibbitts draws on Andrew Harmel-Law's concept of the
Architecture Advice Process --- in which any team can make
any architectural decision as long as it has looked for advice from the affected parties and
documented the decision in an ADR. This approach is presented as a mechanism for replacing permission-seeking
with advice-seeking. The social shift from ``who approves this?'' to ``who should I talk
to before we do this?'' is presented as a cultural unlock that enables decentralized
governance without sacrificing coherence.

% ---------------------------------------------------------------------------
\section{Tool Support for Architectural Decision-Making}
\label{sec:tool-support}

Along with the author's exploratory research, several tools and platforms have been proposed to support architectural decision-making:

\begin{itemize}
  \item \textbf{ADR tools} (e.g., adr-tools, Log4brains) focus on recording and
  versioning Architecture Decision Records (ADRs). They capture the \emph{what} and
  \emph{why} of decisions but do not model uncertainty, impact, or technical plans for
  hypotheses that are not yet decisions. ADRs document settled choices; they are not
  designed to manage the pre-decision state of uncertainty that ArchHypo addresses.

  \item \textbf{Architecture analysis tools} (e.g., ATAM-based workshops, trade-off
  analysis methods) provide structured evaluation of architectural alternatives but are
  typically heavyweight, workshop-based activities rather than continuous,
  team-embedded processes. Their value lies in one-shot or periodic analysis sessions,
  not in the ongoing, iterative management of uncertainty throughout a product's
  lifecycle.

  \item \textbf{Internal Developer Portals} such as Backstage, Port, and Cortex offer
  a catalog-centric view of services and infrastructure. While they improve
  discoverability and documentation, none of these platforms natively supports
  hypothesis engineering or architectural uncertainty management. They provide the
  integration surface that Hypo-Stage exploits, but leave the hypothesis engineering
  layer entirely absent.
\end{itemize}

Hypo-Stage differentiates itself by being explicitly designed around the ArchHypo
lifecycle: from hypothesis elicitation through assessment, technical planning, and
resolution. Its integration with Backstage places this lifecycle within the developer's
daily tooling environment. No existing tool, to the best of the author's knowledge,
provides this combination.

\begin{table}[htbp]
\centering
\caption{Comparison of tools relevant to architectural decision-making support.}
\label{tab:tools-comparison}
\begin{tabular}{p{0.20\linewidth} p{0.16\linewidth} p{0.14\linewidth} p{0.14\linewidth} p{0.14\linewidth} p{0.14\linewidth}}
\toprule
\textbf{Tool / Platform} & \textbf{Primary Purpose} & \textbf{Models Uncertainty} & \textbf{Technical Plans} & \textbf{Lifecycle Tracking} & \textbf{Team-Embedded} \\
\midrule
ADR tools (adr-tools, Log4brains) & Record settled decisions. & No & No & Partial & No \\
\midrule
ATAM / Trade-off workshops & Evaluate architectural alternatives. & Partial & No & No & No \\
\midrule
IDPs (Backstage, Port, Cortex) & Developer portal; catalog; documentation. & No & No & No & Yes \\
\midrule
Hypo-Stage (this work) & Support ArchHypo hypothesis engineering. & Yes & Yes & Yes & Yes (via Backstage) \\
\bottomrule
\end{tabular}
\end{table}

% ---------------------------------------------------------------------------
\section{Empirical Studies of Architectural Practices in Industry}
\label{sec:empirical-studies}

Industry-focused empirical studies on architectural practices remain relatively scarce
compared with tool proposals and method descriptions. Among the most relevant:

\begin{itemize}
  \item The ArchHypo evaluations by ~\cite{silva2024tse,silva2024ieeesw}
  constitute the closest antecedents to this work. They demonstrated the value of
  hypothesis engineering in two distinct settings --- a web-testing startup and a
  mission-critical government budget system --- but relied entirely on manual application
  with no tool support. This work introduces the first evaluation of ArchHypo in a
  tool-mediated setting.

  \item Studies on architectural governance in agile organizations (e.g., the work on
  SAFe and Spotify-model variants) document how companies structure roles and
  ceremonies for architecture, but rarely examine explicit uncertainty-management
  techniques.

  \item The multivocal review conducted during the author's exploratory research surfaced
  practitioner reports (blog posts, conference talks, white papers) describing ad-hoc
  approaches to architectural uncertainty. These reports corroborate the need for
  structured methods but, being gray literature, carry weaker evidential value.
\end{itemize}

Based on the findings of these related studies, this work contributes to the body of empirical evidence by evaluating a
tool-supported application of ArchHypo with multiple teams, addressing a gap in both the ArchHypo literature (which has not yet studied tool-mediated use) and the broader
architectural-practices literature (which lacks studies of hypothesis-engineering tools in scaled settings).
