% ============================================================
%  Chapter 5 — Results and Refinements
%  Master's Thesis: Evaluating and Refining Hypo-Stage
%  José Gonçalves Lima Neto — IME-USP, 2026
% ============================================================
\chapter{Results and Refinements}
\label{chap:results}

This chapter presents the empirical findings of the study and the refinements to
Hypo-Stage that emerged from them. The results are organized around the four
research questions defined in Chapter~\ref{chap:introduction}: RQ1 addresses how
teams used the tool and the technique in practice; RQ2 characterizes the
architectural uncertainties they registered; RQ3 reports the benefits and
challenges perceived by participants; and RQ4 documents each refinement, tracing
it to a concrete observation and to the question it addresses.
Section~\ref{sec:emerging-patterns} goes beyond a direct answer to the four
questions by drawing two families of \textit{emerging patterns}: team-behavior
patterns (Group~A) and a set of novel technical architectural patterns for
hypothesis engineers (Group~B). The latter constitutes a distinct contribution
of this work, addressing a gap left open by prior research in the ArchHypo line,
as discussed in Section~\ref{sec:group-b-patterns}.

% ─────────────────────────────────────────────────────────────
\section{RQ1 — How Teams Used Hypo-Stage in Practice}
\label{sec:rq1-results}
\label{sec:results-rq1}
% ─────────────────────────────────────────────────────────────

\subsection{Hypothesis Registration Dynamics}
\label{subsec:registration-dynamics}

Over the four-week observation period, both teams together registered thirteen
architectural hypotheses in Hypo-Stage. Six hypotheses originated from the
Product~Team and seven from the Platform~Team. This distribution is noteworthy
because the Platform~Team is significantly smaller (four engineers) yet produced
more registered hypotheses than the twelve-engineer Product~Team. This asymmetry
is partly explained by the registration of a dense cluster of five Platform
hypotheses on a single day (12 April 2026), all addressing complementary aspects
of the shared Kafka streaming infrastructure — a burst pattern discussed further
in Section~\ref{sec:group-b-patterns}.

Across both teams, twelve of the thirteen hypotheses were authored by senior
technical leaders — staff or principal engineers who held explicit architectural
responsibilities for their respective domains. One hypothesis was registered by a
mid-level engineer on the Product~Team. Both technical leaders in each team
independently reported that a much larger number of architectural uncertainties
existed in their daily work than the number they managed to register. The
principal constraint they named was time: formulating a well-structured hypothesis
statement requires focused, uninterrupted attention that sprint delivery
pressure rarely afforded. This observation resonates with findings from prior
ArchHypo evaluations, where the learning curve and process-integration demands
were described as significant barriers to adoption
\cite{silva2025archhypo, silva2024archhypo}.

\subsection{Role-Based Engagement}
\label{subsec:role-engagement}

Participant observation and the active listening sessions revealed a clear
stratification in how different roles engaged with Hypo-Stage during the study
period. Senior engineers treated the tool as a decision-support instrument:
they used it to externalize concerns that had until then existed only as tacit
knowledge, and they referenced the hypothesis backlog when discussing
architectural trade-offs in planning sessions. Mid-level and junior engineers
were aware of the tool's existence and attended the training session, but
neither team saw them register hypotheses independently beyond the initial
introduction. This does not indicate disengagement; rather, both technical
leaders explained that junior engineers typically operate at a scope where
architectural uncertainties are resolved quickly — within a sprint or less —
leaving little perceived need for formal registration.

\subsection{Tool Adoption Patterns}
\label{subsec:adoption-patterns}

Usage of Hypo-Stage was non-uniform across the observation window. Registration
activity was concentrated in two periods: a first wave during and immediately
after the training session, and a second, more intensive wave near the end of
the period when the Platform~Team used the tool for a structured architectural
discussion on Kafka failure-handling strategies. Between these waves, both
technical leaders reported using the hypothesis detail view to review existing
records and update technical plans, but the tool was not used in team-visible
ceremonies such as sprint retrospectives or planning meetings. When asked,
participants noted that embedding the tool into these ceremonies would require
an explicit cultural agreement within the team — something that had not yet
occurred within the observation window.

The integration of Hypo-Stage into the Backstage Internal Developer Portal was
cited by both teams as a meaningful enabler of adoption. Because Backstage is
already the daily entry point for service catalog, documentation, and CI/CD
status, engineers encountered Hypo-Stage without navigating to a separate tool.
The EntityHypothesesTab, which surfaces only the hypotheses linked to the
component an engineer is already viewing, was specifically mentioned as reducing
the cognitive cost of context-switching into hypothesis management.

% ─────────────────────────────────────────────────────────────
\section{RQ2 — Architectural Uncertainties Registered}
\label{sec:rq2-results}
\label{sec:results-rq2}
% ─────────────────────────────────────────────────────────────

\subsection{Data Overview}
\label{subsec:data-overview}

Table~\ref{tab:hypotheses-overview} presents the thirteen hypotheses registered
during the study, anonymized in accordance with the ethical agreement described
in Section~\ref{sec:ethical-considerations}. Full exported records (statements,
ratings, entity links, and author notes) appear in Appendix~\ref{app:hypotheses-data}. Each hypothesis is identified by a
short label, its source type (Requirements or Solution), its lifecycle status
at the end of the observation period, and its uncertainty and impact ratings.

\begin{table}[htbp]
\caption{Overview of the thirteen registered architectural hypotheses.
  Req.\ = Requirements-sourced; Sol.\ = Solution-sourced.
  U = Uncertainty level; I = Impact level.
  Ratings: VL = Very Low, M = Medium, H = High, VH = Very High.}
\label{tab:hypotheses-overview}
\centering\small
\begin{tabular}{llllll}
\toprule
\textbf{ID} & \textbf{Short label} & \textbf{Team} & \textbf{Source} &
\textbf{U} & \textbf{I} \\
\midrule
H01 & Kotlin/JOOQ/Spring template readiness & Product  & Req. & VH & VH \\
H02 & Kafka direct exposure via platform framework & Platform & Req. & VL & M  \\
H03 & Decouple eng.\ logs from security/compliance audit & Platform & Req. & M  & VH \\
H04 & Blob storage GDPR right-to-erasure & Product  & Req. & M  & VL \\
H05 & Credit-bureau decoupling seam & Product  & Sol. & H  & VH \\
H06 & Billing aggregation chunked keyset pagination & Product  & Sol. & VL & VH \\
H07 & Blob ownership transfer across service domains & Product  & Req. & VL & VH \\
H08 & Blob cross-client sharing governance & Product  & Req. & VL & VH \\
H09 & DLT for audit vs.\ offset-based replay & Platform & Sol. & M  & VH \\
H10 & Synchronous retry for transient consumer failures & Platform & Sol. & M  & H  \\
H11 & Failure state inside vs.\ outside Kafka flow & Platform & Sol. & M  & H  \\
H12 & Error classification before retry & Platform & Sol. & M  & H  \\
H13 & Payload immutability before re-enqueueing & Platform & Sol. & VL & VH \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Source Types}
\label{subsec:source-types}

Eight of the thirteen hypotheses were classified as \textit{Requirements}-sourced:
the uncertainty stems from an incompletely specified quality-attribute goal or
an external obligation whose implementation path is unknown. Five were classified
as \textit{Solution}-sourced: the team has an existing or candidate architectural
decision and is uncertain about its consequences. The predominance of
requirements-sourced hypotheses on the Product~Team (five out of six) and
the predominance of solution-sourced hypotheses on the Platform~Team's Kafka
cluster (four out of five Kafka hypotheses) reflects the different nature of
each team's work: the Product~Team faces open questions about whether architectural
goals can be met at all, while the Platform~Team is assessing the trade-offs of
specific implementation choices for a shared infrastructure component.

\subsection{Quality Attribute Distribution}
\label{subsec:qa-distribution}

Table~\ref{tab:qa-distribution} summarizes the frequency of quality attributes
across the thirteen hypotheses. Attributes were assigned by the hypothesis authors
using the multi-select form in Hypo-Stage; a hypothesis may carry more than one
attribute.

\begin{table}[htbp]
\caption{Quality attribute frequency across the thirteen hypotheses.}
\label{tab:qa-distribution}
\centering\small
\begin{tabular}{lcc}
\toprule
\textbf{Quality Attribute} & \textbf{Count} & \textbf{\% of hypotheses} \\
\midrule
Reliability        & 6 & 46\% \\
Auditability       & 5 & 38\% \\
Performance        & 4 & 31\% \\
Interoperability   & 4 & 31\% \\
Extensibility      & 3 & 23\% \\
Usability          & 3 & 23\% \\
Scalability        & 2 & 15\% \\
Modifiability      & 2 & 15\% \\
Flexibility        & 2 & 15\% \\
Configurability    & 2 & 15\% \\
Maintainability    & 1 &  8\% \\
Security           & 1 &  8\% \\
Observability      & 1 &  8\% \\
Reusability        & 1 &  8\% \\
\bottomrule
\end{tabular}
\end{table}

Reliability and Auditability emerge as the two dominant concerns. The prominence
of Auditability is especially notable and directly traceable to two contextual
forces: the organization's internal security audit requirements, which drove H03,
and the applicable privacy regulation governing binary document storage (H04),
as well as the Kafka immutability and DLT governance questions (H09, H13).
This concentration reveals a recurring archetype in which external compliance
obligations crystallize into architectural hypotheses — a pattern discussed in
Section~\ref{sec:group-b-patterns} as the \textit{Regulatory Trigger Hypothesis}.

\subsection{Uncertainty and Impact Profile}
\label{subsec:uncertainty-impact}

Nine of the thirteen hypotheses carry a \textit{Very High} or \textit{High}
impact rating, signalling that the teams were not registering peripheral concerns
but uncertainties with potentially significant consequences for schedule, rework,
or system quality. Uncertainty ratings are more evenly distributed: five
hypotheses are rated \textit{Very Low} (the team has strong evidence but the
hypothesis is tracked for governance reasons), four are \textit{Medium}, one
is \textit{High}, and one \textit{Very High}.

The combination of Very Low uncertainty and Very High impact (H06, H07, H08,
H13) is particularly instructive. In ArchHypo's model, this quadrant maps to
hypotheses that \textit{can be postponed} — the team already has sufficient
evidence to act, so the immediate risk is low — yet the downstream consequences
of acting incorrectly are severe. Hypo-Stage's prioritization dashboard surfaced
these hypotheses in the "Can Postpone" segment of its focus panel, enabling
teams to consciously defer them without losing visibility. This represents
a practical instance of the \textit{Postpone Uncertain Decisions} pattern
described in \citet{silva2024patterns}.

\subsection{Technical Plans in Use}
\label{subsec:technical-plans}

Seven of the thirteen hypotheses had at least one associated technical plan at
the end of the observation period. Together these seven hypotheses carried thirteen
individual technical planning entries. The complete catalog of plan titles and
strategies is reproduced in Section~\ref{sec:app-technical-plans} of Appendix~B.
Table~\ref{tab:plan-strategies} summarizes
the strategies used.

\begin{table}[htbp]
\caption{Technical plan strategies observed across the hypothesis data.}
\label{tab:plan-strategies}
\centering\small
\begin{tabular}{lcc}
\toprule
\textbf{Strategy} & \textbf{Plans} & \textbf{Hypotheses using it} \\
\midrule
Guideline Implementation & 3 & 3 \\
Spike (Architectural Spike) & 3 & 2 \\
Modularization / Flexibility & 2 & 1 \\
Experiment & 2 & 2 \\
Tracer Bullet & 1 & 1 \\
Software Analytics (Trigger) & 1 & 1 \\
Other (implementation) & 1 & 1 \\
\midrule
\textbf{Total} & \textbf{13} & — \\
\bottomrule
\end{tabular}
\end{table}

Guideline Implementation and Architectural Spikes were the most frequently
chosen strategies, consistent with findings from the PropSEP/Jetsoft evaluation
\citep{silva2025archhypo}. The Tracer Bullet strategy appeared once — for H10,
where the Platform~Team planned a controlled injection of transient failures into
a real consumer path to measure ordering and latency characteristics. This
application illustrates how the Tracer Bullet differs from a Spike: rather than
an isolated experiment, the plan exercises the full production consumer path
under representative conditions \citep{silva2024agilepatterns}.

Six hypotheses had no technical plan at the end of the study. Three of these
(H01, H04, H02) were registered but not yet acted upon; two (H07, H08) had
been validated through planning actions that predated the observation window and
whose plans were registered retrospectively; and one (H05) was under active
discovery but the plan document was still being prepared.

\subsection{Lifecycle and Status}
\label{subsec:lifecycle-status}

At the end of the observation period, ten hypotheses had \textit{Open} status,
one was \textit{In Review} (H05, the credit-bureau decoupling, which had moved
to a structured discovery phase), and two were \textit{Validated} (H07 and H08,
both in the blob storage domain). No hypothesis was invalidated or discarded
during the study window. The predominance of the Open status is consistent with
the short duration of the study (four weeks, two agile sprints) and with the
nature of the uncertainties: most concern mid- to long-horizon architectural
concerns that teams do not expect to resolve within a single sprint cycle.

% ─────────────────────────────────────────────────────────────
\section{RQ3 — Perceived Benefits and Challenges}
\label{sec:rq3-results}
\label{sec:results-rq3}
% ─────────────────────────────────────────────────────────────

\subsection{Perceived Benefits}
\label{subsec:benefits}

Active listening sessions with all four interviewed participants (the two
technical leaders from each team) converged on three recurring themes regarding
the perceived value of using Hypo-Stage.

\paragraph{Externalization of tacit architectural knowledge.}
Both teams described the act of registering a hypothesis as valuable independent
of what the tool subsequently did with the record. Articulating an uncertainty
in falsifiable form forced a degree of precision that normal sprint conversations
do not require. Participants noted that seeing a hypothesis written down — with
an explicit uncertainty rating, an impact rating, and a related quality attribute
— made it easier to share the concern with colleagues who had not been involved
in the original discussion.

\paragraph{Prioritization support.}
The \textit{Where to Focus} panel of the Hypo-Stage dashboard — distinguishing
hypotheses that \textit{Need Attention} (high uncertainty and high impact) from
those that \textit{Can Postpone} (low impact) — was described by both technical
leaders as a useful aid in architectural conversations. One participant noted that
the "Can Postpone" category in particular gave them a principled way to explain
to product managers why a known concern was being deliberately deferred rather
than ignored.

\paragraph{Collective memory for architectural decisions.}
Both teams observed that the hypothesis backlog served a documentation function
that was distinct from Architecture Decision Records (ADRs) or design documents.
Whereas ADRs capture decisions after they are made, an architectural hypothesis
captures an open question before a decision is available. Participants described
this as filling a gap in their current toolchain: they had good mechanisms for
recording what was decided but poor mechanisms for recording what was still
unknown.

\subsection{Perceived Challenges}
\label{subsec:challenges}

Three challenges were consistently raised across participants.

\paragraph{Negative-formulation learning curve.}
The convention of writing hypotheses as falsifiable statements — phrased in a
form that specifies under which evidence the team would conclude the hypothesis
is incorrect — was described by all four participants as non-intuitive. Engineers
are accustomed to stating what they believe to be true. Inverting that belief
into a falsifiable form that describes what would prove it false requires a
deliberate cognitive effort that is easy to skip under time pressure. This finding
replicates the learning curve reported in both prior ArchHypo evaluations
\citep{silva2025archhypo, silva2024archhypo}.

\paragraph{Registration time cost.}
Both technical leaders described the difficulty of dedicating the focused time
required to register a hypothesis properly. Fast-moving architectural micro-decisions
— resolved within a day or less — were rarely registered because, by the time a
hypothesis could be written, the question had already been answered informally.
This creates a systematic under-representation of short-lifecycle uncertainties
in the hypothesis backlog, which may be consequential when the same question
re-emerges in a future context.

\paragraph{Absence of team-visible ceremonies.}
Neither team had established a recurring ceremony — such as an architectural
review or a hypothesis triage meeting — in which Hypo-Stage records were consulted
as a team. Tool usage remained largely individual during the observation period.
Both technical leaders expressed the view that realizing the collaborative
potential of hypothesis engineering would require the team to institutionalize at
least one touchpoint where the backlog is reviewed together.

% ─────────────────────────────────────────────────────────────
\section{RQ4 — Refinements to Hypo-Stage}
\label{sec:rq4-results}
\label{sec:results-rq4}
\label{sec:refinements}
% ─────────────────────────────────────────────────────────────

This section presents the refinements implemented in Hypo-Stage from commit
\texttt{ceee509} (the artifact boundary established by \citealt{correa2025}) to
commit \texttt{3a53d4e} (HEAD at the time of writing). Each refinement is traced
to a concrete observation from the study and linked to the research question it
primarily addresses. All refinements address RQ4 by definition; selected
refinements also bear on RQ1 (adoption ease) and RQ3 (usability and learning
curve).

\subsection{Backstage Integration and Catalog Surfacing}
\label{subsec:refinement-backstage}

\textbf{Observation.} During the training session and the first week of
observation, both teams navigated to Hypo-Stage via a standalone entry point
that was separate from their regular Backstage workflow. Engineers who were not
technical leaders rarely revisited the tool after the initial session.

\textbf{Refinement.} Hypo-Stage was wired as a first-class Backstage experience,
with routable pages for home, hypothesis creation, detail view, and edit flows.
Crucially, a catalog component tab (\texttt{EntityHypothesesTab}) was introduced,
surfacing all hypotheses linked to the component an engineer is currently viewing
in the Backstage catalog. A feature flag (\texttt{hypo-stage}) was added to allow
gradual, team-controlled rollout.

\textbf{Impact.} Reducing navigation friction lowered the perceived cost of
checking the hypothesis backlog while already working with a service's catalog
entry. The Platform~Team's technical leader noted that the component tab made
hypothesis review a natural side-effect of existing Backstage usage patterns.

\subsection{Backend API and Filtering}
\label{subsec:refinement-api}

\textbf{Observation.} Engineers with cross-team or platform-wide responsibilities
needed to filter hypotheses by team and by specific service component. The initial
version of Hypo-Stage returned all hypotheses without server-side filtering,
which degraded usability as the hypothesis data grew.

\textbf{Refinements.} Three new API capabilities were implemented:
(i)~\texttt{GET /hypotheses} extended with \texttt{entityRef} and \texttt{team}
query parameters; (ii)~\texttt{GET /hypotheses/referenced-entity-refs} to populate
the component filter without listing the entire catalog; and
(iii)~\texttt{GET /hypotheses/teams} deriving team names from catalog metadata
for components referenced by hypotheses.

\subsection{Prioritization Dashboard}
\label{subsec:refinement-dashboard}

\textbf{Observation.} Both technical leaders expressed the desire to quickly
identify which hypotheses deserved immediate attention and which could safely be
deferred. Manually scanning the full hypothesis list imposed an unnecessary
cognitive burden, especially when the backlog grew.

\textbf{Refinement.} A \texttt{HypothesisListDashboard} component was introduced
above the hypothesis table, presenting: (i)~overview cards with total count,
validated count, and hypotheses created in the last thirty days;
(ii)~a \textit{Where to Focus} section distinguishing \textit{Need Attention}
(high uncertainty and high impact) from \textit{Can Postpone} (low impact),
with explanatory subtext aligned with ArchHypo's prioritization model
\citep{silva2025archhypo}; and (iii)~distribution panels for uncertainty and
impact Likert values. The dashboard respects entity, team, and component scoping.

\textbf{Impact (RQ3).} The "Can Postpone" segment was specifically cited by
participants as resolving a communication challenge: it provided a principled
vocabulary for explaining deliberate deferral to non-technical stakeholders.

\subsection{Hypothesis Lifecycle Consistency}
\label{subsec:refinement-lifecycle}

\textbf{Observation.} In the initial implementation, create and update operations
were not wrapped in database transactions, which created a risk of partial writes
(hypothesis row updated without the corresponding event entry, or vice versa).
Artifact analysis revealed two records with missing event entries, suggesting that
this race condition had occurred in practice.

\textbf{Refinements.} Create and update paths were refactored to use explicit
database transactions. Delete was extended to cascade to technical planning rows
and events, with clear API semantics and a confirmation dialog that requires the
exact hypothesis statement before proceeding (paste-enabled to avoid penalizing
users with long statements). Success and error toast notifications were added
for destructive operations.

\subsection{Technical-Planning-Driven Assessment Updates}
\label{subsec:refinement-tp-assessment}

\textbf{Observation.} Engineers reported that after completing a technical plan
action — such as an architectural spike — they updated the hypothesis uncertainty
rating separately from the plan record. This two-step process was error-prone
and led to hypotheses whose assessment was stale relative to their technical
plans.

\textbf{Refinements.} Technical planning create and update operations were
extended to accept optional \texttt{uncertainty} and \texttt{impact} fields.
When present, the service updates the parent hypothesis's Likert values in the
same transaction. The event model was refined to distinguish
\texttt{TECHNICAL\_PLANNING\_CREATE} and \texttt{TECHNICAL\_PLANNING\_UPDATE}
events from generic hypothesis updates, preserving a traceable audit trail of
which assessment changes were driven by plan actions.

\subsection{Evolution Chart Improvements}
\label{subsec:refinement-chart}

\textbf{Observation.} The initial uncertainty/impact evolution chart used opaque
numeric y-axis ticks that participants could not interpret without consulting the
Likert scale documentation separately. Multiple events on the same calendar day
produced ambiguous x-axis labels.

\textbf{Refinements.} The chart was updated to render verbal y-axis ticks
(Very Low through Very High). Event source was made visually distinguishable:
circles for manual or creation-driven points, squares for technical-planning-driven
points. A legend explaining the marker shapes was added. When multiple events
occur on the same calendar day, the x-axis uses date and time to avoid ambiguity.
Tooltips were improved to label technical-planning-driven assessment moves with
the plan number when identifiable.

\subsection{Form Usability Improvements}
\label{subsec:refinement-forms}

\textbf{Observation.} Participants found the hypothesis creation form cognitively
demanding, particularly the uncertainty and impact selectors, which displayed
only numeric values, and the quality attribute field, which accepted free text
without suggesting valid values.

\textbf{Refinements.} Uncertainty and impact selectors were redesigned as Likert
star pickers showing both a numeric button (1–5) and an adjacent text label
(Very Low through Very High) with disambiguating icons. Quality attributes were
replaced by a multi-select autocomplete showing attribute name and description
in the dropdown. Entity references for linking hypotheses to catalog components
were given a catalog-backed autocomplete. Documentation link fields were given an
improved list editor with add/remove and basic URL validation.

\subsection{Summary: Refinements Mapped to Observations}
\label{subsec:refinement-summary}

Table~\ref{tab:refinements-map} summarizes the nine categories of refinements,
the primary observation that motivated each, and the research question each
addresses most directly.

\begin{table}[htbp]
\caption{Refinements to Hypo-Stage mapped to motivating observations and
  research questions. RQ columns indicate primary (P) or secondary (S) linkage.}
\label{tab:refinements-map}
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{p{3.8cm}p{4.5cm}ccc}
\toprule
\textbf{Refinement} & \textbf{Motivating observation} &
\textbf{RQ1} & \textbf{RQ4} & \textbf{RQ3} \\
\midrule
Backstage catalog tab
  & Infrequent revisit by non-leaders outside native workflow
  & S & P & S \\
Backend API filtering
  & Cross-team users needed scoped views as hypothesis data grew
  & S & P & — \\
Prioritization dashboard
  & Manual scanning of full backlog was cognitively costly
  & — & P & S \\
Lifecycle transaction consistency
  & Partial writes produced stale event records
  & — & P & — \\
TP-driven assessment updates
  & Assessment and plan records drifted apart post-spike
  & — & P & S \\
Evolution chart improvements
  & Numeric y-axis and same-day collisions caused confusion
  & — & P & S \\
Form usability improvements
  & Numeric-only selectors; free-text QA field caused errors
  & — & P & P \\
Safer deletion UX
  & Long statements made confirmation hostile without paste
  & — & P & — \\
Toast notifications
  & No feedback after destructive operations caused re-submission
  & — & P & — \\
\bottomrule
\end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────
\section{Emerging Patterns}
\label{sec:emerging-patterns}
% ─────────────────────────────────────────────────────────────

Beyond the direct answers to RQ1–RQ4, thematic analysis of observation notes,
active listening transcripts, and the combined hypothesis data produced two
families of patterns. Group~A describes team-behavior conditions that shape
whether hypothesis engineering becomes a sustainable practice. Group~B describes
recurring structural archetypes in the content of the architectural hypotheses
themselves. These two families address orthogonal dimensions: Group~A asks
\textit{when} hypothesis engineering works; Group~B asks \textit{what shape}
hypotheses take in practice.

% ─────────────────────────────────────────────────────────────
\subsection{Group A — Team Behaviour Patterns for Hypothesis Engineering
            Adoption}
\label{sec:group-a-patterns}
% ─────────────────────────────────────────────────────────────

The four Group~A patterns emerge from triangulating RQ1 and RQ3 findings. They
describe organizational and individual-level conditions that the study observed
to facilitate or obstruct the adoption of hypothesis engineering as a team
practice. These patterns complement the existing ArchHypo pattern languages
\citep{silva2024patterns, silva2023initialpatterns}, which focus on the
mechanics of hypothesis assessment and deferral, by addressing the social and
operational context in which those mechanics must operate.

\paragraph{A1 — Senior-Curated Hypothesis Backlog.}
\textbf{Problem.} How can a team ensure that mid- and long-horizon architectural
uncertainties remain visible and actively managed over time, even though
individual engineers work at different scopes and timescales?

\textbf{Observed evidence.} Twelve of the thirteen registered hypotheses were
authored by senior technical leaders — engineers who carried explicit
architectural responsibility and whose work horizon extended beyond the current
sprint. Non-senior engineers did not sustain independent hypothesis registration
after the initial training episode. The hypotheses that were registered shared
a common trait: they concerned uncertainties whose consequences would materialize
weeks or months in the future, making them invisible to typical sprint-scope
review practices.

\textbf{Solution.} Assign senior engineers as the primary curators of the
hypothesis backlog. Their architectural horizon and organizational authority
position them to recognize which uncertainties are worth formalizing. Non-senior
participation should be enabled by the tool but cannot be assumed to be
self-sustaining in the early phases of adoption.

\paragraph{A2 — Fast-Resolution Blind Spot.}
\textbf{Problem.} Many architectural micro-decisions are made and resolved within
hours or a single day. Their speed makes them feel too transient to register, yet
their cumulative architectural effect may only become visible — as drift or
inconsistency — long after the original decision context is forgotten.

\textbf{Observed evidence.} Both technical leaders reported that the majority of
architectural decisions encountered day-to-day were resolved before a hypothesis
could be written. This speed made registration feel retrospectively unnecessary.
Yet both also described recurring situations where the same question re-surfaced
in a new context, requiring the team to reconstruct reasoning that had previously
been resolved informally.

\textbf{Solution.} Introduce a lightweight periodic review — for example, a brief
end-of-sprint ceremony — specifically to identify decisions made in the past
iteration that resolved quickly but may carry cumulative consequences. Decisions
with cross-component impact or quality-attribute implications are candidates for
retroactive registration or lightweight decision records.

\paragraph{A3 — Registration as a Slack-Dependent Practice.}
\textbf{Problem.} Translating a tacit architectural concern into a properly
formulated ArchHypo hypothesis requires focused, uninterrupted attention.
Sprint delivery pressure makes this kind of attention scarce and unreliable as
a basis for sustained practice.

\textbf{Observed evidence.} Senior leaders explicitly described the difficulty
of "dedicating proper time to register all possible hypotheses they encounter
day-to-day." The observed set of thirteen hypotheses is known to
under-represent the total uncertainties that leaders were aware of. Registration
activity clustered around protected discussion time — the training session and
a structured team session on Kafka strategy — rather than being distributed
across normal sprint work.

\textbf{Solution.} Treat hypothesis registration as a deliberate activity
requiring protected time — a recurring architectural hygiene session, separate
from sprint ceremonies, where engineers externalize concerns that have been
accumulating in tacit knowledge. The IDP integration of Hypo-Stage reduces
navigation cost but does not substitute for protected cognitive time.

\paragraph{A4 — Negative-Formulation Barrier.}
\textbf{Problem.} The ArchHypo convention of writing hypotheses as falsifiable
statements — specifying the evidence that would prove the hypothesis false — is
cognitively inverted relative to how engineers normally articulate beliefs.
This inversion is a persistent adoption barrier even for practitioners who
understand the technique conceptually.

\textbf{Observed evidence.} All four interviewed participants mentioned that the
falsifiability requirement is non-intuitive and difficult to apply under time
pressure. This replicates findings reported in the PropSEP/Jetsoft evaluation
\citep{silva2025archhypo} and remains unresolved in the current version of
the technique's guidance materials.

\textbf{Solution.} Provide statement templates that scaffold the inversion:
starting from an architectural belief ("We believe X is true") and guiding the
engineer to express what observable outcome would prove it false. Embedding this
scaffolding inline in the hypothesis creation form — as contextual guidance
adjacent to the statement field — would reduce the cognitive overhead at the
moment of registration without requiring prior training.

% ─────────────────────────────────────────────────────────────
\subsection{Group B — Technical Architectural Patterns for Hypothesis
            Engineers}
\label{sec:group-b-patterns}
% ─────────────────────────────────────────────────────────────

The prior work in the ArchHypo research line has established the technique's
core concepts \citep{silva2025archhypo}, documented patterns for hypothesis
assessment, uncertainty management, and decision deferral
\citep{silva2024patterns, silva2023initialpatterns}, and reported on the
technique's learning curve and process-integration challenges
\citep{silva2024archhypo}. What this body of work has not yet addressed is
a question that only becomes answerable through empirical data on
hypotheses registered in practice:

\begin{quote}
\textit{What recurring shapes do architectural hypotheses take when teams apply
ArchHypo to real systems? And what structural properties of a hypothesis — its
source type, quality-attribute profile, and relation to its organizational
context — should an engineer recognize in order to write and manage it more
effectively?}
\end{quote}

This gap was also identified in calls for future work within the ArchHypo
pattern literature, which explicitly noted the need for "identifying and
documenting patterns for various types of hypotheses" and for "exploration of
common uncertainties related to specific quality attributes"
\citep{silva2024patterns}. The thirteen-hypothesis dataset produced by this
study constitutes the first empirical dataset that enables such identification.
The five patterns described below are therefore a \textbf{novel contribution}
of this work: they are grounded in real hypothesis data, not derived
from theoretical analysis, and they provide a vocabulary that hypothesis
engineers can use both to recognize familiar situations more quickly and to
communicate about hypothesis types across teams.

\paragraph{B1 — Hypothesis Cluster.}
\textbf{Context.} A team is responsible for a shared platform component on which
multiple other service teams depend. An architectural question about that
component cannot be decomposed into a single hypothesis because the design space
has several interrelated dimensions, each of which involves a distinct trade-off.

\textbf{Problem.} Formulating a single hypothesis for a multi-dimensional
architectural concern either oversimplifies the problem (losing important
trade-off structure) or produces an unmanageably broad statement that is hard
to falsify.

\textbf{Observed evidence.} Five hypotheses (H09–H13) all concern the shared
Kafka streaming infrastructure and were registered on the same day, as the
Platform~Team's technical leader used Hypo-Stage to structure a design discussion.
The five hypotheses address: (i)~DLT audit semantics versus offset-based replay;
(ii)~synchronous versus asynchronous retry for transient consumer failures;
(iii)~failure-state placement inside versus outside the Kafka flow;
(iv)~error classification before retry; and (v)~payload immutability before
re-enqueueing. Each is independently falsifiable yet together they map the full
decision space for consumer-side failure handling in the shared streaming
platform.

\textbf{Solution.} When a shared platform component has architectural uncertainty
involving multiple interrelated dimensions, decompose it into a cluster of
related but independently falsifiable hypotheses. Assign the cluster to the same
platform component in the tool so it appears together in the component-scoped
tab. Manage technical plans at the individual hypothesis level but review the
cluster as a unit in architectural discussions, as the resolution of one
hypothesis may constrain the options available for others.

\paragraph{B2 — Regulatory Trigger Hypothesis.}
\textbf{Context.} A software system operates in a regulated domain — financial
services, healthcare, data privacy — where external compliance obligations
create architectural constraints whose implementation path is not fully known.

\textbf{Problem.} A compliance requirement imposes an architectural obligation
before the team has determined how to satisfy it. It is unclear whether the
existing architecture already satisfies the obligation, what changes would be
required if it does not, and what the cost of non-compliance would be.

\textbf{Observed evidence.} Two hypotheses in the dataset arise directly from
regulatory obligations: H04 (the GDPR right-to-erasure requirement for binary
documents in the blob storage service) and H03 (organization-wide security
audit requirements for database log correlation across the platform framework).
In both cases the uncertainty is not about whether to comply — that is a legal
given — but about whether the current architecture \textit{can} comply and at
what technical cost.

\textbf{Solution.} When a regulatory obligation creates an architectural
constraint, formulate it as a hypothesis stating what the system must be capable
of and making falsifiable the belief that the current architecture satisfies this
capability. This transforms compliance work from an implicit assumption into an
explicit tracked uncertainty with a technical plan and a validation target. The
quality attributes Auditability and Security are diagnostic markers for this
pattern.

\paragraph{B3 — Decoupling Seam Hypothesis.}
\textbf{Context.} A service currently handles multiple responsibilities — domain
logic, external integrations, and identity management — in a tightly coupled
implementation. The team believes the service could be restructured to serve
multiple domain contexts, but the boundaries of the decoupling and the migration
path are unknown.

\textbf{Problem.} Deciding \textit{whether} to decouple is itself uncertain: the
team does not know what the correct contract boundary should be, which existing
flows can be migrated without business-logic changes, and what the downstream
impact on other teams would be. Committing to decoupling without this information
risks a disruptive refactor that delivers less reuse than expected.

\textbf{Observed evidence.} H05 (credit-bureau decoupling) captures this
precisely. The service currently handles identity, bureau integration, and
confidential document flows using domain-specific entities tied to a single
business vertical. The hypothesis asks whether these integrations can be
decoupled into context-agnostic middleware accepting only each provider's minimum
required identifiers, reusable across domains without duplicated cost. At the
end of the observation period, H05 was the only hypothesis in \textit{In Review}
status, with a discovery technical plan mapping (i)~the dependency matrix of
current versus real dependencies per integration, (ii)~an agnostic interface
contract proposal, and (iii)~a migration plan identifying which flows can isolate
immediately.

\textbf{Solution.} When a team suspects that a service boundary is too narrow
for the reuse it should enable, formulate the uncertainty as a decoupling seam
hypothesis targeting the specific boundary in question. The technical plan should
follow a discovery strategy before any implementation is attempted: map existing
dependencies, propose an interface contract free of domain-specific concepts,
and identify the migration sequence. Quality attributes Interoperability,
Extensibility, and Reusability are diagnostic markers for this pattern.

\paragraph{B4 — Load-Validated Hypothesis.}
\textbf{Context.} A team has proposed a specific algorithmic or implementation
approach to handle a high-volume data processing challenge. The team has already
conducted a capacity or volume test at scale, producing evidence that the
approach works. However, the hypothesis has not been formally closed because
the implementation is still in progress.

\textbf{Problem.} The hypothesis occupies the lowest uncertainty quadrant
(Very Low) while retaining the highest impact rating (Very High), reflecting a
situation where the team \textit{knows} the solution works but the architectural
commitment is still consequential. The hypothesis must be tracked until
implementation is complete, yet its characterization as "Open" may suggest
unresolved doubt that no longer exists.

\textbf{Observed evidence.} H06 (billing aggregation chunked keyset pagination)
exemplifies this. The team validated the approach — chunked keyset pagination
with isolated repository-level transactions — through a capacity test at five
times production scale (45 million rows total, 1.5 million rows per month).
Profiling confirmed that JPA stream processing caused first-level cache growth,
and that chunking scopes transactions to approximately ten thousand rows so the
garbage collector can reclaim memory, producing a healthy sawtooth heap pattern.
The hypothesis remains Open because the implementation is ongoing, but the
uncertainty has been effectively resolved.

\textbf{Solution.} When a team has strong empirical evidence that a solution
approach is correct but implementation is incomplete, retain the hypothesis in
Open status with a Very Low uncertainty rating and document the validation
evidence in the hypothesis notes. Plan the hypothesis lifecycle so that it
transitions to Validated upon completion of the production implementation, not
upon completion of the test. The hypothesis note field should reference the
capacity test artifact explicitly, preserving the evidentiary chain.

\paragraph{B5 — Shared Resource Governance Hypothesis.}
\textbf{Context.} A platform service manages a shared resource — binary objects,
credentials, configuration data — on which multiple product domains depend. New
business requirements demand that the resource be shared across domain boundaries
or transferred between owning services, but the governance rules for these
operations have not been established.

\textbf{Problem.} The team must decide how to model cross-domain access to the
shared resource — specifically, which service "owns" the resource at any given
point, who may read or write it, and under what conditions ownership can be
transferred. These questions have significant security, consistency, and usability
implications. Without explicit governance decisions, different product teams will
adopt incompatible access patterns, creating consistency violations that are
difficult to detect and remedy.

\textbf{Observed evidence.} H07 (blob ownership transfer across service domains)
and H08 (blob cross-client sharing governance) are two instances of this pattern
in the secure blob storage domain. H08 was driven by a concrete usage scenario
in which a customer-facing channel receives documents that are later routed to
support tooling; the support team needs read access without breaking the blob
vault's ownership invariants. Both hypotheses were validated through a Guideline
technical plan — an alignment session that produced a governance decision enabling
prioritization of the transfer and sharing implementation work. Quality attributes
Security, Interoperability, and Usability are diagnostic markers for this pattern.

% ─────────────────────────────────────────────────────────────
\section{Revisiting the Research Questions}
\label{sec:rq-revisit}
% ─────────────────────────────────────────────────────────────

This section concludes the analytical body of the chapter by stating explicitly how
the preceding empirical and design material bears on the four research questions
formulated in Chapter~\ref{chap:introduction}. The intent is \emph{consolidative}:
no additional empirical claims are introduced; the aim is to make the logical
organization of the chapter legible relative to those questions.

\subsection{Correspondence between research questions and chapter sections}

Table~\ref{tab:rq-chapter-map} summarizes where each research question is principally
addressed in this chapter, and where integrated interpretation (Group~A/B) extends
the principal sections.

\begin{table}[htbp]
\centering
\small
\setlength{\tabcolsep}{5pt}
\caption{Correspondence between the research questions (Chapter~\ref{chap:introduction})
  and the organization of evidence in this chapter.}
\label{tab:rq-chapter-map}
\begin{tabular}{p{0.12\linewidth} p{0.40\linewidth} p{0.40\linewidth}}
\toprule
\textbf{RQ} & \textbf{Principal analysis} & \textbf{Integrated interpretation} \\
\midrule
RQ1 & Section~\ref{sec:rq1-results} (tool use and adoption dynamics). &
Section~\ref{sec:group-a-patterns} (Group~A; read together with RQ3). \\
\midrule
RQ2 & Section~\ref{sec:rq2-results} (typological account of the data). &
Section~\ref{sec:group-b-patterns} (Group~B structural archetypes). \\
\midrule
RQ3 & Section~\ref{sec:rq3-results} (benefits and challenges). &
Section~\ref{sec:group-a-patterns} (Group~A; read together with RQ1). \\
\midrule
RQ4 & Section~\ref{sec:rq4-results} (nine refinement categories, observation-led). &
--- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Brief interpretive synthesis}

For RQ2, the question couples a typological obligation with a structural one.
Section~\ref{sec:rq2-results} discharges the former by characterizing what was
registered; Section~\ref{sec:group-b-patterns} discharges the latter by articulating
recurrent structural forms in the data. Jointly, they satisfy the formulation of
RQ2 adopted in Chapter~\ref{chap:introduction}.

For RQ1 and RQ3, the Group~A patterns provide an integrated reading: they condense
conditions for sustaining hypothesis engineering that are visible only when
observational evidence on use (RQ1) and on perceived benefits and friction (RQ3) is
examined together. The wording of RQ1 and RQ3 is not revised here; the contribution
is to clarify what an adequate empirical answer must cover in the present setting.

For RQ4, Section~\ref{sec:rq4-results} reports refinements in the manner expected
in design-science accounts of artifact evolution, that is, as outcomes grounded in
observations rather than as a separate implementation chronicle.
