%!TeX root=../thesis.tex

\chapter{Conclusion}
\label{ch:conclusion}
\label{chap:conclusion}

The central question motivating this dissertation was whether embedding the ArchHypo
technique into a tool deployed within a team's existing workflow could lower the barriers
that prior evaluations had identified as threats to its adoption. Both prior studies of
ArchHypo --- at the Catch Solve startup \citep{silva2024ieeesw} and at the PropSEP
mission-critical project \citep{silva2024tse} --- demonstrated the technique's conceptual
value, yet neither involved a supporting tool or a multi-team professional software development setting,
and both documented a consistent limitation: without tool support and expert
facilitation, hypothesis engineering remained a fragile, senior-anchored practice. This
work takes up that limitation directly. By deploying Hypo-Stage as a Backstage plugin
within a large-scale financial services organization and observing two teams over approximately nine calendar weeks of real sprint work (five consecutive two-week sprints between the first and last hypothesis registration dates in Hypo-Stage; Chapter~\ref{chap:research-design}), the study produces the first empirical account of ArchHypo in
a tool-mediated setting --- and in doing so, extends the empirical base of a research line
whose prior publications have called explicitly for this kind of evidence
\citep{silva2024patterns}.

The results both confirm and complicate the expectations established by prior work.
On the side of confirmation: hypothesis registration remained predominantly senior-led
across both teams, the falsifiability requirement continued to be described as cognitively
demanding under sprint pressure, and the absence of team-visible ceremonies for reviewing
the hypothesis backlog emerged as a structural obstacle --- all consistent with findings
from the PropSEP evaluation \citep{silva2024tse}. On the side of complication: the
Backstage integration proved to be a genuinely meaningful enabler rather than merely a
convenience. The EntityHypothesesTab, which surfaces hypotheses in the context of the
service an engineer is already viewing, reduced context-switching cost in a way that
participants named explicitly as a design improvement over standalone tools. This
suggests that the integration surface matters as much as the hypothesis engineering
mechanics themselves --- a finding that is absent from prior ArchHypo literature because
no prior study involved a tool.

\section{Synthesis of Contributions}
\label{sec:conclusion-synthesis}

Taken together, the four research questions produce an account of how hypothesis
engineering behaves when it is embedded in a professional team's daily tooling, rather
than applied as an expert-facilitated intervention.

With respect to \textbf{how teams used Hypo-Stage in practice} (RQ1), the study
reveals a pattern of senior-anchored, burst-mode adoption. Both technical leaders
treated the hypothesis backlog as an externalization instrument for concerns that had
until then existed only as tacit knowledge; they used registration events as structured
prompts for architectural discussions --- most visibly during the Platform Team's
Kafka failure-handling session --- rather than as a continuous logging practice. This
usage pattern carries a significant implication: the value of Hypo-Stage in the observed
context was less about individual hypothesis tracking and more about creating
structured occasions for architectural reasoning. Junior and mid-level engineers
remained peripheral to registration, not because of the tool's interface but because
their day-to-day work horizon rarely extended beyond sprint scope. This observation
connects to the broader literature on tacit knowledge in architectural decision-making
\citep{souza2019}: the knowledge asymmetry that prevents less experienced engineers
from participating meaningfully in architectural discussions is not dissolved by tool
availability alone.

With respect to \textbf{the structural character of the architectural uncertainties
registered} (RQ2), the thirteen-hypothesis dataset reveals a landscape dominated by
compliance-driven and shared-infrastructure concerns. Reliability and Auditability
emerged as the dominant quality attributes, reflecting the operational reality of a
scaled financial services product subject to both internal security audit and
data-privacy regulation. The five Group~B archetypes --- Hypothesis Cluster, Regulatory
Trigger Hypothesis, Decoupling Seam Hypothesis, Load-Validated Hypothesis, and
Shared Resource Governance Hypothesis --- constitute the first empirically grounded
vocabulary for the recurring shapes that architectural hypotheses take when ArchHypo
is applied in a professional software development context. Their significance lies not
in frequency claims --- the dataset is too small for that --- but in naming: by giving
recognizable structure to hypothesis types that engineers encounter intuitively but
cannot currently articulate, the archetypes provide a vocabulary that can accelerate
both hypothesis formulation and cross-team communication. The relationship between
these archetypes and the existing ArchHypo pattern language
\citep{silva2024patterns} is one of complementarity: the existing patterns address
the mechanics of assessment and deferral; the Group~B archetypes address the content
of what is being assessed.

With respect to \textbf{perceived benefits and challenges} (RQ3), the study
documents a consistent tension between the intellectual value that participants
attributed to hypothesis management and the operational cost they experienced in
exercising it. The formalization of tacit architectural knowledge into a traceable,
prioritized artifact was perceived as genuinely valuable --- participants described
hypotheses as filling a gap that ADRs leave open, capturing the pre-decision state of
uncertainty rather than the post-decision rationale. Yet this value was consistently
held hostage to sprint delivery pressure and to the cognitive demand of the
falsifiability requirement. The four Group~A organizational patterns (A1--A4)
condense these tensions into actionable propositions: they suggest that the adoption
of hypothesis engineering as a sustainable team practice depends not only on tool
usability but on the organizational conditions --- protected time, explicit role
assignment, statement scaffolding, and ceremonial touchpoints --- that the tool alone
cannot provide.

With respect to \textbf{empirically grounded refinements to Hypo-Stage} (RQ4), the
nine refinement categories produced during the observation period operationalize a
formative evaluation cycle in which observations feed directly into tool changes that
are then redeployed and re-observed. This cycle is consistent with the action-research
element of the research design and demonstrates that the formative evaluation model
advocated in flexible-design guidance \citep{robson2016} is well-suited to tool
artifacts whose value proposition is tightly coupled to workflow fit. The most
structurally significant refinement --- the evolution chart that tracks uncertainty and
impact ratings across successive assessments --- directly addresses the absence of
assessment history in the initial Hypo-Stage version \citep{correa2025} and provides
teams with a mechanism for evaluating whether their technical plan actions are having
the intended effect. That this refinement emerged from observation rather than from
design foresight is itself a finding: the value of longitudinal assessment tracking
only becomes visible when the tool is used in a real product context over multiple sprint cycles.

\section{Implications for Practice and Research}
\label{sec:implications}

The findings of this study carry implications at two levels: for practitioners who are
considering adopting hypothesis-engineering tools in their organizations, and for
researchers who study tool-mediated architectural practices.

For \textbf{practitioners}, the most actionable implication concerns the integration
surface of the tool. The study demonstrates that embedding Hypo-Stage within Backstage
--- the platform where engineers already manage services, documentation, and CI/CD
pipelines --- removed a class of adoption friction that a standalone tool would have
created. This is consistent with the broader platform engineering literature on Internal
Developer Portals \citep{correa2025} and with the Team Topologies argument that tooling
should follow, rather than reshape, team cognitive pathways \citep{skelton2019}.
A general principle suggested by these findings is that hypothesis-engineering tools
are more likely to achieve sustainable adoption when they are delivered as extensions
to an existing workflow rather than as a new workflow to be added to an already busy
sprint cadence.

A second implication for practitioners concerns the organizational conditions for
broad participation. Tool availability is a necessary but not sufficient condition for
distributing architectural decision-making beyond senior engineers. The evidence from
this study --- that twelve of thirteen hypotheses were authored by staff or principal
engineers even with tool support in place --- suggests that organizations intending to
use ArchHypo to operationalize the socio-technical architecture paradigm described in
\citep{infoq2024trends} should combine tool deployment with three complementary investments:
protected time for architectural hygiene sessions separate from sprint ceremonies
(Group~A pattern A3); statement scaffolding embedded in the hypothesis creation form
to lower the falsifiability barrier (Group~A pattern A4, actualized in the Hypo-Stage
form improvements); and at least one recurring team-visible ceremony where the
hypothesis backlog is reviewed collectively. Without these conditions, the tool risks
reinforcing the senior-anchor dependency it was designed to dissolve.

For \textbf{researchers}, the study offers three methodological observations that may
benefit future work in tool-mediated empirical software engineering. First, the
formative evaluation design --- in which refinements are implemented during the
observation period and their effects are observed within the same study --- provides
richer insight into the causal relationship between tool design choices and adoption
behavior than a retrospective evaluation would. This design is especially appropriate
for artifacts like Hypo-Stage, whose value proposition is tightly coupled to usability
and workflow integration. Second, participant observation combined with artifact
analysis of the tool's own data (the hypothesis records) proved a productive combination:
the field notes captured social dynamics that the artifact data could not, while the
artifact data provided a concrete, verifiable record against which observational
interpretations could be tested. Third, even a short observation window --- approximately nine calendar weeks (five two-week sprints spanned by registrations),
thirteen hypotheses --- is sufficient to identify initial structural patterns in hypothesis
content, provided the observation is conducted in a naturalistic organizational setting
rather than a controlled experiment. This supports the view, consistent with
\citet{robson2016}, that flexible case-study designs can produce theoretically
generative findings even when the dataset is limited in scale.

More broadly, this study contributes to the conversation about what it means to
support distributed architectural decision-making with tools. The vision of a
socio-technical architecture \citep{infoq2024trends, infoq2025empowering, infoq2025decentralized} in which
architecture work is shared across teams rather than centralized in a dedicated function
depends on practitioners at all levels having both the vocabulary and the tooling to
surface, assess, and communicate architectural uncertainties. Hypo-Stage, as evaluated
here, represents one operational instantiation of that vision. The findings suggest
that the tool's contribution is genuine but partial: it succeeds in lowering
context-switching cost and in making hypothesis management available within the daily
workflow, but the deeper challenge --- widening participation beyond senior engineers ---
remains an organizational and pedagogical problem that no tool, by itself, can solve.

\section{Limitations and Future Work}
\label{sec:future-work}
\label{sec:conclusion-limitations}

Any interpretation of the findings presented in this dissertation must be read against
its methodological boundaries. Acknowledging these boundaries is not a disclaimer but
a research responsibility: it makes explicit the conditions under which the findings hold
and identifies the empirical work required to extend them.

\subsection{Scope and Transferability}

The study was conducted within a single technology organization that already operated
a mature Backstage deployment and maintained a scaled financial services product in
Latin America. The enabling condition of Backstage availability is non-trivial: it means
that findings about adoption friction --- specifically, that the EntityHypothesesTab reduced
context-switching cost --- may not transfer directly to organizations without an Internal
Developer Portal. Similarly, the dominance of Reliability and Auditability as quality
attributes in the hypothesis dataset reflects the specific regulatory environment of the
study site; organizations in other domains may exhibit very different hypothesis profiles.
The organizational culture, the team size differential between the Product and Platform
teams, and the presence of a researcher who was also the tool's supervising author
all constitute contextual specifics whose influence cannot be fully separated from the
findings. These characteristics do not invalidate the results; they define the population
to which the results most directly apply and orient the replication agenda described
below.

\subsection{Dataset Scope and Hypothesis Lifecycle}

Thirteen hypotheses produced by two teams over approximately nine calendar weeks (five two-week sprints from first to last registration) is a dataset appropriate for
initial conceptual characterization but insufficient for statistical or frequency claims.
The Group~A organizational patterns and Group~B structural archetypes should be read
as emerging propositions --- grounded in real empirical data and internally consistent
with the theoretical framework --- rather than as consolidated findings. Their value at
this stage is nominative: they provide a vocabulary for recognition that can be tested and
refined in future studies. That observation window was also too short to observe
full hypothesis lifecycle completion: eleven of the thirteen hypotheses remained in Open
status at the study's end, and the tool's ability to support teams through validation,
invalidation, and retrospective learning after hypothesis closure remains empirically
unexamined.

\subsection{Researcher Role and Potential Bias}

The researcher's dual position as organizational employee and academic investigator
introduces a known potential for bias. Although strategies were employed to maintain
role separation --- including explicit disclosure to participants of the academic framing,
collection of field notes distinguishing observational from interpretive entries, and
submission of the coding scheme to supervisory review --- the researcher's status as
the tool's supervising author and as a domain expert in ArchHypo may have influenced
how participants described their experience. Social desirability effects cannot be ruled
out, and the absence of a comparison condition means that it is not possible to
distinguish the effect of the tool from the effect of the structured technique itself, or
from the Hawthorne effect of researcher presence.

\subsection{Future Work Agenda}

The most immediate priority is \textbf{replication in additional organizational contexts},
including organizations without an existing Backstage deployment, with teams of varying
size, domain, and tenure profile, and ideally with a researcher who does not hold a
pre-existing relationship with the tool. Replication studies could allow the Group~A and
Group~B patterns to be tested against new hypothesis datasets, providing the basis for
either consolidating them into the ArchHypo pattern language \citep{silva2024patterns}
or revising them in light of disconfirming evidence.

A \textbf{longitudinal design} spanning multiple development cycles within the same
organization would address the lifecycle limitation directly. Such a study could allow
observation of hypothesis closure events --- the moment at which a hypothesis is
validated or invalidated and a corresponding architectural decision is taken or deferred
--- and of whether the tool adequately supports retrospective learning once the
hypothesis is closed. It would also permit examination of the longer-term trajectory of
senior-anchor dependency: whether, over time and with repeated practice, hypothesis
registration becomes more broadly distributed across experience levels, or whether the
senior-anchored pattern is structurally stable.

The \textbf{integration of Hypo-Stage with Architecture Decision Record tools}
such as Log4brains represents a natural architectural extension of the research. Hypo-Stage
manages the pre-decision lifecycle of architectural uncertainty; ADR tools manage the
post-decision record. Connecting the two would create a single traceable thread from
hypothesis registration through technical plan execution to final architectural decision,
addressing a gap in the current tool landscape identified in Chapter~3. Empirical
investigation of whether this integration changes how teams reason about the
relationship between uncertainty management and decision documentation would
constitute a meaningful contribution to the architectural knowledge management literature.

Finally, the \textbf{question of broadening participation} beyond senior engineers
remains the most consequential open problem in the ArchHypo adoption research agenda.
The findings of this study, and of the two prior ArchHypo evaluations, consistently
point to a pattern in which the technique's conceptual value is recognized but its
practice remains concentrated in the hands of those who already possess architectural
fluency. Intervention studies that test whether guided hypothesis templates, asynchronous
review workflows, or explicit architectural hygiene ceremonies can measurably increase
registration breadth --- and that measure the quality, not only the volume, of the
resulting hypotheses --- would provide the empirical basis for translating the Group~A
organizational patterns from emerging propositions into actionable adoption guidelines.

Beyond the four strategic directions above, the nine emerging patterns summarized
in Table~\ref{tab:emerging-patterns} open a concrete, pattern-anchored research
agenda. Chapter~\ref{chap:results} deliberately leaves each pattern
in descriptive form, without prescribing a solution; the studies outlined below
sketch the empirical work that would be required before any such prescription
could be justified.

For the Group~A patterns (team-behavior conditions for sustainable hypothesis
engineering):

\begin{itemize}
  \item \textbf{A1 --- Senior-Curated Hypothesis Backlog:} investigate
    longitudinally whether explicit curation roles and permission or review
    models can shift authorship distribution across experience levels without
    degrading hypothesis quality.
  \item \textbf{A2 --- Fast-Resolution Blind Spot:} design and evaluate a
    lightweight end-of-sprint or retroactive registration workflow that captures
    architectural micro-decisions whose cumulative effect only becomes visible
    after the fact.
  \item \textbf{A3 --- Registration as a Slack-Dependent Practice:} compare
    sustained registration rate and hypothesis quality under (i)~a periodic
    architectural-hygiene cadence with protected cognitive time versus (ii)~in-sprint
    ad~hoc registration.
  \item \textbf{A4 --- Negative-Formulation Barrier:} design and test statement
    template scaffolding --- for instance, inline contextual guidance embedded
    in the hypothesis creation form --- to measure whether falsifiability
    correctness improves without prior training.
\end{itemize}

For the Group~B patterns (structural archetypes of architectural hypotheses):

\begin{itemize}
  \item \textbf{B1 --- Hypothesis Cluster:} investigate decomposition strategies
    and cluster-level review practices for multi-dimensional platform concerns,
    and evaluate how cluster-aware tooling affects traceability and decision
    velocity.
  \item \textbf{B2 --- Regulatory Trigger Hypothesis:} evaluate whether modeling
    compliance obligations as tracked architectural hypotheses, with explicit
    falsification criteria against the current architecture, improves regulated-domain
    outcomes compared with implicit compliance assumptions.
  \item \textbf{B3 --- Decoupling Seam Hypothesis:} evaluate a discovery-first
    technical-plan pipeline (dependency mapping, agnostic interface design,
    staged migration) against ad~hoc refactoring, measuring disruption and
    realized reuse.
  \item \textbf{B4 --- Load-Validated Hypothesis:} define and empirically test
    lifecycle states that distinguish ``empirically validated but not yet
    implemented'' from ``Open with unresolved uncertainty,'' and evaluate
    reviewer comprehension across the two.
  \item \textbf{B5 --- Shared Resource Governance Hypothesis:} investigate how
    governance rules (ownership, cross-domain read and write, transfer) for
    shared platform resources can be expressed and tracked as architectural
    hypotheses, and measure the downstream effect on consistency violations.
\end{itemize}
